---
title: "toponym"
date: January 24, 2024
output:
  github_document
  
fontsize: 12pt
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

Welcome to the `toponym` GitHub page!

The `toponym` package supplies users of R with tools to visualize and analyze toponym (= place name) distributions. It is intended as an interface to the [GeoNames](https://www.geonames.org/) data. A regular expression filters data and a map is created displaying locations which comply with it. The functions make data and plots available for further analysis -- either within R or in the working directory. Users can select regions within countries, provide coordinates to define regions, or specify a region within the package to restrict the data selection to that region or compare regions with the remainder of countries.

## Installation
In order to install this package, you will need `devtools`. You can download and load the current development version of `toponym` from [GitHub](https://github.com/Lennart05/toponym) with:

``` r
# install.packages("devtools")
# library ("devtools")
devtools::install_github("Lennart05/toponym")
```

## Create a simple map

The function `top()`, meaning "toponym", creates maps of places complying with a regular expression. Minimally one or more strings and one or more countries (in that order) are given as input. The following code is a simple example of this:

```{r example}
library(toponym) # load the package
top("itz$", "DE")
```

The plot displays all locations which end in "-itz" in Germany, their total frequency (2182), and stores the data in the global environment.

## Country designations

The data is meant to cover maps and toponyms of the world. The function `country()` lets users access all permitted country and region designations used by this package. The query `country table` returns the entire data frame.

```{r}
head(country(query = "country table"))
```
If you want to access the row of a specific country, you can either provide the ISO2 code, ISO3 code or the country name:

```{r country data}
country(query = "Argentina")
# returns the respective row for Argentina

country(query = "ARG")
# returns the same row
```

If `regions` is set to `1`, the function returns all region designations:

```{r}
country("Mali", regions = 1)
# returns all region names and IDs of Mali available in the data
```



## Frequent toponym substrings

`topFreq()` lets users find strings frequently recurring in toponym. A simple example for the Philippines would be:
```{r}
topFreq(countries = "Philippines",
        len = 3,
        limit = 10,
        type = "$")
```
Among all toponyms in the data for the Philippines (`countries = "Philippines"`), these are the ten (`limit = 10`) most frequent trailing (`type = "$"`) strings consisting of (a length of) three characters (`len = 3`).

The additional parameter `polygon` allows users to restrict the data to a subset of the selected countries. Only toponyms within the polygon are selected. The polygon must intersect a country specified by the parameter `countries`. The package contains a predefined polygon for the historical Danelaw area of England for purposes of illustration: 

```{r}
topFreq(countries = "GB",
        len = 3,
        limit = 10,
        polygon = toponym::danelaw_polygon
)
```
## Create polygons

Coordinates which delimit a polygon are input in the form of a data frame. The `createPolygon()` function helps users to define their own polygon by point-and-click or to retrieve map data.
```{r eval=FALSE}
argentina_polygon <- createPolygon(countries = "AR", regions = 1)
```
In this example, a map of Argentina `AR` with highest-level administrative borders `regions = 1` will appear as a plot. Now users can click to set points which define a polygon. The last point should not repeat the first point. In RGui, users exit the point selection by middle-clicking or right-clicking and then pressing stop. In RStudio, users exit the point selection by pressing ESC or Finish in the top right corner of the plot. Once finished, a data frame with longitudinal and latitudinal coordinates called `argentina_polygon` is created. 


## Strings specific to a region

`topComp()` determines which toponym strings in the data are characteristic to a region. Consider again the following example for the Danelaw area:

```{r list of candidates}
topComp(countries = "GB",
       len = 3,
       limit = 100,
       rat = .8, 
       polygon = toponym::danelaw_polygon
       )
```
The function compares the frequency of trailing strings (`type = "$"`) within the Danelaw area (`polygon = toponym::danelaw_polygon`) with their frequency in the United Kingdom (`countries = "GB"`) and returns a data frame. The output is in descending order by their proportional frequency. The search is limited to the 100 (`limit = 100`) most frequent strings in theUnited Kingdom consisting of (a length of) three characters (`len = 3`). The cut-off ratio of 80% (`rat = .8`) means that at least 80% of all occurrences (in the country or countries) must be inside the polygon. In this case, the string "-rpe" occurs 175 times in the United Kingdom and 156 of these 176 occurrences are within the target polygon resulting in a ratio percentage of 89.14%.

## Creating multiple maps at once

```{r multiple maps, eval = FALSE}
topCompOut(countries = "GB",
           len = 3,
           limit = 75,
           rat = .8,
           polygon = toponym::danelaw_polygon
                )
```

Running this with the same settings as just used for `topComp()` produces a distributional map and a data frame of every string. The plots are saved in the working directory in a separate folder called "plots". The data frames are saved in another folder called "data frames".


## Apply a Z-test 

`topZtest()` tests whether the frequency of a toponym string is significantly greater in the given area than in the rest of the country or countries:

```{r ztest flanders}
topZtest(strings = "aat$",
           countries = "BEL",
           polygon = toponym::flanders_polygon
                )
```
In this example, the function compares the toponymic distribution of the trailing string "-aat" (`strings = "aat$"`) in Flanders (`polygon = toponym::flanders_polygon`) with Belgium (`countries = "BEL"`) as a whole. The result of the two proportion test is returned as an object of class `htest`.

## The functions
The core functions are as follows:

* `top()` returns and plots selected toponyms onto a map.
* `country()` helps in navigating designations of countries and regions used by the package.
* `creatPolygon()` lets users create a polygon by point-and-click or directly retrieve polygon data.
* `topComp()` compares toponyms in a polygon and in the remainder of a country (or countries).
* `topCompOut()` saves multiple maps and toponym data.
* `topFreq()` retrieves most frequent toponyms.
* `topZtest()` lets users apply a Z-test on toponym distributions.

For help type `?toponym` or a question mark following the individual function name (or use the `help()` syntax). A link to the index at the bottom of each help page provides a useful way of navigating the package.

## Regular expression

For a concise description of which regular expressions exist and how they can be used, type `help("regex")` in the R console or follow [this guide](https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html).

## Data

The toponym data comes from [GeoNames](https://www.geonames.org/) and will be automatically downloaded when you call any of the core functions. It is recommended to save the data of the countries you access in the package directory. This is the default option of the function `getData()` but it is possible to place it in the temporary folder by changing the parameter to `save = FALSE`. If you want to store data only temporary, you need to use `getData()` before any other function. Type `tempdir()` to find the temporary directory of the current session.
For mapping purposes as well as region designations, the [geodata](https://cran.r-project.org/web/packages/geodata/index.html) package is used. It provides spatial data for all countries and regions available in this package. All maps are stored in the geodata package directory.
