---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Toponym

<!-- badges: start -->
<!-- badges: end -->

The initial goal of the `toponym` package is having a tool to easily visualize the distribution of toponyms. When a regular expression is given, a map is generated displaying all populated locations matching it. If needed, the plot can be saved locally as .png file. A list of locations in form of a data frame can be temporary saved in the session environment or to the working directory as .csv file as well.

For questions about which toponyms are prevalent in a region, the `toponym` package searches for the most frequent toponym prefixes or suffixes, checks how often they appear in said region, and generates maps of those toponyms which significantly frequently appear there. This way, you will quickly have a large number of maps helping you finding toponyms which show distributions that are fruitful to be examined further.

## Installation
In order to install this package, you will need `devtools`. You can download and load the current development version of `toponym` from [GitHub](https://github.com/Lennart05/toponym) with:

``` r
# install.packages("devtools")
# library ("devtools")
devtools::install_github("Lennart05/toponym")
```

## Create a simple map

The function `top()`, meaning "toponym", creates maps of places matching the regular expression. By default, it only looks at places in Germany, as the script was initially created to find slavic toponyms in Germany. The following code is a simple example of this:

```{r example}
library(toponym) # load package
top("itz$") 

```

The plot shows all locations, which end with "-itz" in Germany, their total frequency, and stores the data in the global environment. As you can see, most occurrences are located in the former slavic settlement zone, indicating that the ending may be of Slavic origin. In this case, we know that already.

## List endings specific to a region

If we want to find out, which toponyms appear frequently in one region compared to the rest, we can run this:

```{r list of candidates}
top.candidates(countries = "DE", count = 75, len = 3, rat = .8)
```

The output is a data frame giving us information about the ratio and frequency of each ending. This means essentially that 95.78% of the places ending with "-itz" are in the polygon roughly covering the former slavic settlement zone. Even though the ending is less common in total, "-kow" may be of interest too as all occurrences are found in the polygon. To be clear what happened: Other common endings such as "-orf", most of which coming from "-dorf", are filtered out since they did not match our fairly high threshold of 80% `rat = .8`. `count = 75` and `len = 3` means that we filtered the 75 most frequent endings with a three character length. 

Instead of applying this data frame with interesting candidates on your own, there is another function doing this automatically. 


## Creating multiple maps at once

```{r multiple maps, eval = FALSE}
candidates.maps(countries = "DE", count = 75, len = 3, rat = .8)
```

Running this with the same settings leaves us with a map of every ending from before saved in the working directory in a separate folder called "plots" as well as the respective data frames in another folder called "data frames". At this point, you can skim through or closely examine the maps at a later date.

## Exemplary usage
To illustrate `toponym`, we make a quick step by step rundown. Let's look at Belgium: The two main regions, Flanders and Wallonia, have each one dominant standard language, alongside dialectal variations. Dutch is dominant in the northern region, Flanders, and French is dominant in the southern region, Wallonia, except small German communities near the German border. Is this division also reflected by toponyms? 
First, we need at least a polygon of one region. As part of the package, the coordinates for both regions are provided as `toponym::flanders_polygon` and `toponym::wallonia_polygon`. Thus, we could run this:
```{r}
### find endings typical in Flanders
head(top.candidates(countries = "BE", count = 100, len = 3, rat = 0.8, type = "$", lons = toponym::flanders_polygon$lons, lats = toponym::flanders_polygon$lats))

```
As the data frame shows there are many endings almost only appearing in Flanders but the limitation to three characters could cut the actual endings off. For example, let's take a closer look at "-aat" which is the second most frequent in total:
```{r}
top("aat$", "BE")

```
In the middle region the higher density of places ending with "-aat" stands out and as we knew from the data frame earlier almost all (99.71%) appear in our polygon. Looking at the full names (after generating this map we end up with a data frame of the selection in the global environment), it's notable that the ending is in fact "-straat" meaning as much as "-street". From here on, a closer examination could start, e. g., asking which conditions led to the frequency in the middle region. And in addition, we could again use the `candidates.maps()` function with the same parameters and check each map:

```{r eval=FALSE}
candidates.maps(countries = "BE", count = 100, len = 3, rat = 0.8, type = "$", lons = toponym::flanders_polygon$lons, lats = toponym::flanders_polygon$lats)
```
## Finding prefixes
As one last example we look at dominant prefixes in the other region Wallonia:
```{r}
head(top.candidates("BE", 100, 4, 0.8, "^", toponym::wallonia_polygon$lons, toponym::wallonia_polygon$lats))

```

Next, we could look at one of them to see their distribution and form. We suspect that it is of French origin and include France and Luxembourg:



```{r}

top("^Gran", c("BE", "FR", "LU"))

```
Most cases seem to be at the border and some, but very few, appear in Flanders too. The great majority stems from "Grand" or "Grande" -- an exception is for example "Grange". As we might have expected, "Gran-" occurs also throughout France but curiously not in Luxembourg. In order to eliminate candidates such as "Grange", the length parameter "len" must be increased and the function needs to be run multiple times with different lengths. 



## the functions
The core functions are again as follows:

* `top()` generates one map with all locations matching the regular expression 
* `top.candidates()` generates a list of prefixes or suffixes frequent in one given region (as polygon)
* `candidates.maps()` generates maps and lists on your computer out of these frequent prefixes and suffixes

For help type `?function` 

## Limitations

It should be clear that this tool merely helps one find frequent toponyms. Less frequent ones, ones with different spelling yet of the same origin or others which are of multiple origins but have the same form (e.g. "-au" in Germany) can only be found by in depth examination.

## Regular expression

The possible queries with regular expressions are broader and wider than these examples illustrate. For a concise description of which regular expressions exist and how they can be used, head over to [the guide](https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html) provided by CRAN.

## Data

The data is kindly provided by [GeoNames](https://www.geonames.org/) and will be automatically downloaded from there. It is recommended to save the data of the countries you look at in the package directory. This is the default option of the function `get.data()` but it is possible to place it in the temporary folder by changing the parameter to `save = FALSE`.

## Polygon

The polygons were created with [Google My Maps](https://www.google.com/maps/about/mymaps/). The functions allow you to provide your own polygon if you want to test other regions. This is a provisional solution and is planned to be replaced by a service within R. First, this requires users to have a Google account and second, the polygon creation and export is not adapted in such a way that it's easily and quickly possible. 



