---
title: "toponym"
date: October 24, 2023
output:
  github_document
  
fontsize: 12pt
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## toponym

<!-- badges: start -->
<!-- badges: end -->

Welcome to the `toponym` GitHub page!

The initial goal of the `toponym` package is to have a tool to easily visualize the distribution of toponyms (= place names). When a regular expression is given, a map is generated displaying all populated locations matching it. If needed, the plot can be saved locally as .png file. A list of locations in form of a data frame can be temporarily saved in the session environment or to the working directory as .csv file as well.

For questions about which toponyms are prevalent in a region, the `toponym` package searches for the most frequent toponym prefixes or suffixes, checks how often they appear in the specified region, and generates maps of those toponyms having significant frequency. This way, you will quickly have a large number of maps helping you to find toponyms which show distributions that may fruitfully be examined further.

## Installation
In order to install this package, you will need `devtools`. You can download and load the current development version of `toponym` from [GitHub](https://github.com/Lennart05/toponym) with:

``` r
# install.packages("devtools")
# library ("devtools")
devtools::install_github("Lennart05/toponym")
```

## Create a simple map

The function `top()`, meaning "toponym", creates maps of places matching a regular expression. First, a string, then a country is given. The following code is a simple example of this:

```{r example}
library(toponym)
top("itz$", "DE")
```

The plot shows all locations, which end with "-itz" in Germany, their total frequency (2182), and stores the data in the global environment. As you can see, most occurrences are located in the former Slavic settlement zone, indicating that the ending may be of Slavic origin. In this case, we know that already.

## Country names and codes

The data is supposed to cover maps and toponyms of the world. The `country.data()` lets you access the table with all countries available as well as regional names of specified countries. 

```{r}
head(country.data(query = "country table"))
```
If you want to access the row of one specific country, you can either provide the ISO2 code, ISO3 code or the country name:

```{r country data}
country.data(query = "Argentina")
# returns the respective row for Argentina

country.data(query = "ARG")
# returns the same row
```

When `regions` is set to `TRUE`, the function returns all region names:

```{r}
country.data(query = "AR", regions = TRUE)
# returns all regional names of Argentina in the data set
```


## List suffixes specific to a region

If we want to find out which toponyms appear frequently in one region compared to the rest of the given countries, we need to specify a polygon. The function requires two vectors, one with the longitudes and another with latitudes. A few polygons are part of the package. Later, the built-in function to create polygons with is described. A polygon covering an area of the Danelaw is in `toponym::danelaw_polygon`. For example, we can run this:

```{r list of candidates}
top.candidates(countries = "GB", count = 75, len = 3, rat = .8, lons = toponym::danelaw_polygon$lons, lats = toponym::danelaw_polygon$lats)
```
The output is a data frame giving us information about the ratio and frequency of each ending. This means essentially that 94.86% of the German places ending with "-rpe" are in the polygon. Even though the endings are less common in total, "-sb" and "-rby" may be of interest too as most occurrences are found in the polygon. To be clear what happened: Other common suffixes in Great Britian such as "-ton" are filtered out since they did not match our fairly high threshold of 80% `rat = .8`. `count = 75` and `len = 3` means that we filtered the 75 most frequent suffixes with a length of three-characters. 

Instead of applying this data frame with interesting candidates on your own, there is another function doing this automatically. 

## Creating multiple maps at once

```{r multiple maps, eval = FALSE}
candidates.maps(countries = "GB", count = 75, len = 3, rat = .8, lons = toponym::danelaw_polygon$lons, lats = toponym::danelaw_polygon$lats)
```

Running this with the same settings leaves us with a map of every ending from before saved in the working directory in a separate folder called "plots" as well as the respective data frames in another folder called "data frames". At this point, you can skim through or closely examine the maps at a later date.

## Create polygons

You can either provide the coordinates of any polygon you have or create one with the built-in function `create.polygon()`. 
```{r eval=FALSE}
argentina_polygon <- create.polygon(countries = "AR", regions = 1)
```
In this example, a map of Argentina `AR` with state boundaries `regions = 1` appears as plot. Now, a polygon can be defined by clicking on the map and pressing `ESC` or the middle mouse button to exit. The last point should not repeat the first point. Once finished, the newly defined polygon is stored as a data frame called `argentina_polygon`. If you do not store the output, the function will only print out the data frame with the coordinates. 
It's also possible to specify names of regions in the given countries. Check the `country.data()` section above for more information.

## Exemplary usage
To illustrate `toponym`, we make a quick step by step rundown. Let's look at Belgium: The two main regions, Flanders and Wallonia (we ignore Brussels), have each one dominant standard language, alongside dialectal variations. Dutch is dominant in the northern region, Flanders, and French is dominant in the southern region, Wallonia, except small German communities near the German border. Is this division also reflected by toponyms? 
First, we need at least one polygon of a region. As part of the package, the coordinates for Flanders is provided as `toponym::flanders_polygon`. Thus, we could run this:
```{r}
### find suffixes typical in Flanders
head(top.candidates(
  countries = "BE", count = 100, len = 3, rat = 0.8,
  lons = toponym::flanders_polygon$lons, lats = toponym::flanders_polygon$lats
))
```
As the data frame shows, there are many suffixes almost only appearing in Flanders but the limitation to three characters could cut the actual suffixes off. For example, let's take a closer look at "-aat" which is the second most frequent in total:
```{r}
top("aat$", "BE")
```
The high density of places ending with "-aat" in the central region stands out and as we knew from the data frame earlier almost all (99.71%) appear in our polygon. Looking at the full names (after generating this map we end up with a data frame of the selection in the global environment), it's notable that the ending is in fact "-straat" meaning as much as "-street". From here on, a closer examination could start, e. g., asking which conditions led to the frequency in the middle region. And in addition, we could again use the `candidates.maps()` function with the same parameters and check each map:

```{r eval=FALSE}
candidates.maps(
  countries = "BE", count = 100, len = 3, rat = 0.8,
  lons = toponym::flanders_polygon$lons, lats = toponym::flanders_polygon$lats
)
```
## Finding prefixes
As one last example we look at dominant prefixes in the same region. For that the parameter `type` needs to be set to "^" as the default option searches for suffixes ("$"):
```{r}
head(top.candidates("BE", 100, 4, 0.8,
  type = "^",
  toponym::flanders_polygon$lons, toponym::flanders_polygon$lats
))
```

Next, we could look at one of them to see their distribution and form. We suspect that it is of Germanic origin and include Netherlands and France for testing:

```{r}
top("^Hoog", c("BE", "NL", "FR"))
```
Most cases seem to be in the western and only few in the central region near the border. As we might have expected, "Hoog-" occurs also throughout the Netherlands but not in France. It is advised that the user adjusts the length parameter `len` and runs the function multiple times to skim through possible toponyms. 


## Frequent toponyms

Lastly, `top.freq()` lets users find the most frequent toponyms in the given countries or a polygon within the countries. A simple example would be:
```{r}
top.freq("Philippines", len = 3, count = 10)
```
You need to specify the number of frequent toponyms you want (here `count = 10`) and the length of the endings or beginning (here `len = 3`). It's also possible to restrict it to one polygon. In comparison to the previous functions, this only outputs what is most frequent in the region or country, while `top.candidates()` calculates, which toponyms are most frequent in the country and then checks which of them meet the threshold specified by `rat`, indicating that they are common to the polygon. Running this function and the previous one we used for the Danelaw again:

```{r}
top.freq("GB",
  lons = toponym::danelaw_polygon$lons,
  lats = toponym::danelaw_polygon$lats, len = 3, count = 10
)
```
```{r}
top.candidates(
  countries = "GB", count = 75, len = 3, rat = .8,
  lons = toponym::danelaw_polygon$lons,
  lats = toponym::danelaw_polygon$lats
)
```

We see that "-rpe" is among the top ten most common endings but "-ton" is more frequent in total. It may be common across the entire country.

## The functions
The core functions are again as follows:

* `top()` generates one map with all locations matching the regular expression
* `country.data()` helps in navigating references to countries (ISO-codes) and regions
* `create.polygon()` lets the user define a polygon by clicking on a map
* `top.candidates()` generates a list of prefixes or suffixes frequent in a given region
* `candidates.maps()` generates maps and lists on your computer out of these frequent prefixes and suffixes
* `top.freq()` generates a list of the most frequent toponyms
* `z.test()` lets users apply a z-test

For help type `?function` or `?toponym`

## Limitations

It should be clear that this tool merely helps one find frequent toponyms. Less frequent ones, ones with different spelling yet of the same origin or others which are of multiple origins but have the same form (e.g. "-au" in Germany) require in depth examination. The data may be non-exhaustive and could contain errors.


## Regular expression

The possible queries with regular expressions are broader and wider than the given examples illustrate. For a concise description of which regular expressions exist and how they can be used, type `help("regex")` in the R console or follow [this guide](https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html).

## Data

The data comes from [GeoNames](https://www.geonames.org/) and will be automatically downloaded from there when you call any of the core functions. It is recommended to save the data of the countries you look at in the package directory. This is the default option of the function `get.data()` but it is possible to place it in the temporary folder by changing the parameter to `save = FALSE`. Type `tempdir()` to find the temporary directory of the current session.

## Future work

We plan to update and improve the package in the future. This includes a more detailed documentation, feedback for users when errors occur and more. We look forward to any feedback or ideas!


